---
title: "SDMs MHS"
format: pdf
execute:
  echo: false
editor: visual
---

## Example of application

Considering only two species to see if its worthwhile:

1.  `r worrms::wm_id2name(217352)`
2.  `r worrms::wm_id2name(217426)`

Load maps and points:

```{r}
suppressPackageStartupMessages(library(terra))
suppressPackageStartupMessages(library(arrow))
terraOptions(progress=0)

species_a <- rast("results/sdms/taxonid=217352/model=mhs/predictions/taxonid=217352_model=mhs_method=maxnet_cog.tif")
species_b <- rast("results/sdms/taxonid=217426/model=mhs/predictions/taxonid=217426_model=mhs_method=maxnet_cog.tif")

sp_a_pts <- read_parquet("results/sdms/taxonid=217352/model=mhs/taxonid=217352_model=mhs_what=fitocc.parquet")
sp_b_pts <- read_parquet("results/sdms/taxonid=217426/model=mhs/taxonid=217426_model=mhs_what=fitocc.parquet")

par(mfrow=c(1,2))
plot(species_a[[1]], range = c(0,1));points(sp_a_pts, pch = 20, cex = .5)
plot(species_b[[1]], range = c(0,1));points(sp_b_pts, pch = 20, cex = .5)
```

Apply masks to isolate to "native" areas:

```{r}
sp_a_mask <- rast("results/sdms/taxonid=217352/model=mhs/predictions/taxonid=217352_model=mhs_mask_cog.tif")
sp_b_mask <- rast("results/sdms/taxonid=217426/model=mhs/predictions/taxonid=217426_model=mhs_mask_cog.tif")

species_a <- mask(species_a, sp_a_mask$native_ecoregions)
species_b <- mask(species_b, sp_b_mask$native_ecoregions)

par(mfrow=c(1,2))
plot(species_a[[1]], range = c(0,1));points(sp_a_pts, pch = 20, cex = .5)
plot(species_b[[1]], range = c(0,1));points(sp_b_pts, pch = 20, cex = .5)
```

Convert to binary format (maybe we can do without it also - in fact, it may be a better choice, but interpretation is a little bit more tricky):

```{r}
sp_a_th <- read_parquet("results/sdms/taxonid=217352/model=mhs/metrics/taxonid=217352_model=mhs_what=thresholds.parquet")
sp_b_th <- read_parquet("results/sdms/taxonid=217426/model=mhs/metrics/taxonid=217426_model=mhs_what=thresholds.parquet")

species_a[species_a < mean(sp_a_th$p10[sp_a_th$model == "maxent"])] <- 0
species_b[species_b < mean(sp_b_th$p10[sp_b_th$model == "maxent"])] <- 0

species_a[species_a > 0] <- 1
species_b[species_b > 0] <- 1

par(mfrow=c(1,2))
plot(species_a[[1]], range = c(0,1));points(sp_a_pts, pch = 20, cex = .5)
plot(species_b[[1]], range = c(0,1));points(sp_b_pts, pch = 20, cex = .5)
```

Sum results to get a "richness" indicator:

```{r}
multi_sp <- sum(species_a, species_b, na.rm = T)

par(mfrow = c(1,2))
plot(multi_sp[[1]], main = "Current period")
plot(multi_sp$ssp245_dec50, main = "SSP2 (2050)")
```

Once we aggregate multiple species, we can assess how much richness will change in the WHSs.
